Thesis Proposal

### 2022-03-18 proposal/proposal.tex
finalize proposal

\paragraph{}
Shared data structures have become an essential field in distributed algorithms research.
We are reaching the physical limits of how many transistors we can place on a CPU core. The industry solution to provide more computational power is to increase the number of cores of the CPU. This is why distributed algorithms have become important. It is not hard to see why multiple processes cannot update sequential data structures designed for one process. For example, consider two processes trying to insert some values into a sequential linked list simultaneously. Processes $p,q$ read the same tail node, $p$ changes the next pointer of the tail node to its new node and after that $q$ does the same. In this run, $p$'s update is overwritten. One solution is to use locks; whenever a process wants to do an update or query on a data structure, the process locks it, and others cannot use it until the lock is released. Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing. Moreover, locks do not allow complete parallelism since only the one process holding the lock can make progress. Our approach is to create a lock-free queue data structure that developers can safely use without the disadvantages of locks. Also there is a common problem in previous shared queues we're going to resolve which make their worst case slow.

The question that may arise is, ``What properties matter for a lock-free data structure?''.
Since executions on a shared data structure are different from sequential ones, the correctness conditions also differ. To prove a concurrent object works perfectly, we have to show it satisfies safety and progress conditions. A \textit{safety condition} tells us that the data structure does not return wrong responses, and a \textit{progress property} requires that operations eventually terminate.

@@

There are various progress properties; the strongest is wait-freedom, and the more common is lock-freedom. An algorithm is \textit{wait-free} if each operation terminates after a finite number of its own steps. We call an algorithm \textit{lock-free} if, after a sufficient number of steps, one operation terminates. A wait-free algorithm is also lock-free but not vice versa since a lock-free algorithm may terminate after an exponential number of steps.
%From now on, we are going to focus on wait freedom.

@@

\section{Related Work}
\subsection{List based queues}
In the following paragraphs, we look at previous lock-free queues.
Michael and Scott~\cite{DBLP:conf/podc/MichaelS96} introduced a lock-free queue which we refer to as the MS-queue. A version of it is included in the standard Java Concurrency Package. Their idea is to store the queue elements in a singly-linked list (see Figure~\ref{fig::msq}). Head points to the first node in the linked list that has not been dequeued, and Tail points to the last element in the queue. To insert a node into the linked list, they use atomic primitive operations like \texttt{LL/SC} or \texttt{CAS}. If $p$ processes try to enqueue simultaneously, only one can succeed, and the others have to retry. This makes the amortized number of steps to be $\Omega(p)$ per enqueue. Similarly, dequeue can take $\Omega(p)$ steps.

\begin{figure}[t]
  \center\includegraphics[scale=0.5]{pics/msqueue}
  \caption{\label{fig::msq}MS-queue structure, enqueue and dequeue operations. Red arrow shows new pointer and gray dashed arrow shows the old arrow.}
\end{figure}

@@

\begin{figure}[t]
  \center\includegraphics[scale=0.3]{pics/baskets}
  \caption{Baskets queue idea.}
\end{figure}

@@

\begin{figure}[t]
  \center\includegraphics[scale=0.5]{pics/sundell}
    \caption{\label{fig::sundell}Global pointer point to arrays. Head and Tail elements are blue dequeued elements are red and current elements of the queue are in green.}
\end{figure}

@@

%Nikolaev and Ravindran~\cite{DBLP:journals/corr/abs-2201-02179} wCQ to be completed.
In the worst-case step complexity of all the list-based queues discussed above, there is a $p$ term that comes from the case all $p$ processes try to do an enqueue simultaneously. Morrison and Afek call this the \textit{CAS retry problem}~\cite{DBLP:conf/ppopp/MorrisonA13}. It is not limited to list-based queues and array-based queues share CAS retry problem as well~\cite{DBLP:conf/spaa/TsigasZ01,DBLP:conf/icdcn/Shafiei09,DBLP:conf/iceccs/ColvinG05} . We are focusing on seeing if we can implement a queue in sublinear steps in terms of $p$ or not.

\subsection{Universal constructions}
Herlihy discussed the possibility of implementing shared objects from other objects~\cite{10.1145/114005.102808}. A \textit{universal construction} is an object that can implement a shared object from any given sequential object. We can implement a concurrent queue using a universal construction. Jayanti proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions~\cite{DBLP:conf/podc/Jayanti98a}. He also introduced a construction that achieves $\textsc{O}(\log^2 p)$ shared accesses~\cite{DBLP:conf/podc/ChandraJT98}. His universal construction can be used to create any data structure, but its implementation is not practical because of using unreasonably large-sized \texttt{CAS} operations.

Ellen and Woelfel introduced an implementation of a Fetch\&Inc object with step complexity of log $p$ using $O(\log n$)bit \texttt{LL/SC} objects, where $n$ is the number of operations~\cite{10.1007/978-3-642-41527-2_20}. Their idea has similarities to Jayanti's construction, and they represent the value of the Fetch\&Inc using the history of successful operations.

@@

\section{Our Contribution}
In this work, we design a queue with $O(\log^2 p +\log q)$ steps per operation, where $q$ is the size of the queue, outperforming queues built by the LinkedList approach. Our idea is similar to Jayanti and Petrovic's multi-enqueuer single-dequeuer Queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but we do not use \texttt{CAS} operations with big words and do not put a limit on the number of concurrent dequeues. We introduce a data structure that allows processes to agree on the linearization ordering of their operations using $\textsc{O}(\log p)$ \texttt{CAS} per operation called a \textit{block tree}. Then we use the block tree as a stepping stone towards our queue algorithm.
A block tree is a tournament tree shared among $p$ processes (see Figure~\ref{fig::blocktree}). Each process has a leaf, and it appends its operations to its leaf. After that, the process tries to propagate its new operation up to the tree's root. An ordering of operations propagated up to a node is stored in that node. All processes agree on the sequence stored in the root and this is used as the linearization ordering.

@@

We also implement methods \texttt{Get(i)}, \texttt{Index(op)} to get the $i$th propagated operation and compute the rank of a propagated operation in the linearization. \texttt{Get(i)} finds the block containing the $i$th operation in the root and then finds its sub-block recursively to reach a leaf. \texttt{Index()} is  similar but more complicated, finding super-blocks from a leaf to the root. The main challenge in each level of \texttt{Get(i)} and \texttt{Index(op)} is that it should take polylogarithmic steps with respect to $p$. After appending operations to the root, processes can find out information about the linearization ordering using \texttt{Get(i)} and \texttt{Index(op)}. Now we can implement Enqueue and Dequeue using our block tree. An \texttt{Enqueue(e)} appends an operation with input argument \texttt{e} in the block tree. To do a \texttt{Dequeue()}, first process $p$ first appends a \texttt{DEQ} operation to the tree. Then $p$ finds the rank of the \texttt{DEQ} using \texttt{Index()}, using the rank of the \texttt{DEQ} and the information stored in the root about the queue $p$ computes the rank of  the \texttt{ENQ} having the answer of the \texttt{DEQ}. Finally $p$ returns the argument of that \texttt{ENQ} using \texttt{Get(i)}.

>>>>

\paragraph{}
Shared data structures have become an essential field in distributed algorithms research.
We are reaching the physical limits of how many transistors we can place on a CPU core. The industry solution to provide more computational power is to increase the number of cores of the CPU. This is why distributed algorithms have become important. It is not hard to see why multiple processes cannot update sequential data structures designed for one process. For example, consider two processes trying to insert some values into a sequential linked list simultaneously. Processes $p,q$ read the same tail node, $p$ changes the next pointer of the tail node to its new node and after that $q$ does the same. In this run, $p$'s update is overwritten. One solution is to use locks; whenever a process wants to do an update or query on a data structure, the process locks it, and others cannot use it until the lock is released. Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing. Moreover, locks do not allow complete parallelism since only the one process holding the lock can make progress. Our approach is to create a lock-free queue data structure that developers can safely use without the disadvantages of locks. Also we are going to resolve a common problem in previous shared queues which makes their worst case slow.

The question that may arise is, ``What properties matter for a lock-free data structure?'', since executions on a shared data structure are different from sequential ones, the correctness conditions also differ. To prove a concurrent object works perfectly, we have to show it satisfies safety and progress conditions. A \textit{safety condition} tells us that the data structure does not return wrong responses, and a \textit{progress property} requires that operations eventually terminate.

@@

There are various progress properties; the strongest is wait-freedom, and the more common is lock-freedom. An algorithm is \textit{wait-free} if each operation terminates after a finite number of its own steps. We call an algorithm \textit{lock-free} if, after a sufficient number of steps, one operation terminates. A wait-free algorithm is also lock-free but not vice versa, in an infinite run of a lock-free algorithm there might be an operation that makes progress every often but never terminates.
%From now on, we are going to focus on wait freedom.

@@

\section{Related Work}
\subsection{List-based Queues}
In the following paragraphs, we look at previous lock-free queues.
Michael and Scott~\cite{DBLP:conf/podc/MichaelS96} introduced a lock-free queue which we refer to as the MS-queue. A version of it is included in the standard Java Concurrency Package. Their idea is to store the queue elements in a singly-linked list (see Figure~\ref{fig::msq}). Head points to the first node in the linked list that has not been dequeued, and Tail points to the last element in the queue. To insert a node into the linked list, they use atomic primitive operations like \texttt{LL/SC} or \texttt{CAS}. If $p$ processes try to enqueue simultaneously, only one can succeed, and the others have to retry. This makes the amortized number of steps to be $\Omega(p)$ per enqueue. Similarly, dequeue can take $\Omega(p)$ steps.

\begin{figure}[t]
  \center\includegraphics[scale=0.5]{pics/msqueue}
  \caption{\label{fig::msq}MS-queue structure, enqueue and dequeue operations. Red arrows show new pointers and gray dashed arrows show the old pointers.}
\end{figure}

@@

\begin{figure}[t]
  \center\includegraphics[scale=0.3]{pics/baskets}
  \caption{Baskets queue idea. There is a mechanism in the algorithm that orders the operations in a basket by the order of which \texttt{CAS}es first successfully.}
\end{figure}

@@

\begin{figure}[t]
  \center\includegraphics[scale=0.5]{pics/sundell}
    \caption{\label{fig::sundell}Global pointers point to arrays. Head and Tail elements are blue, dequeued elements are red and current elements of the queue are green.}
\end{figure}

@@

%Nikolaev and Ravindran~\cite{DBLP:journals/corr/abs-2201-02179} wCQ to be completed.
In the worst-case step complexity of all the list-based queues discussed above, there is a $p$ term that comes from the case all $p$ processes try to do an enqueue simultaneously. Morrison and Afek call this the \textit{CAS retry problem}~\cite{DBLP:conf/ppopp/MorrisonA13}. It is not limited to list-based queues and array-based queues share the CAS retry problem as well~\cite{DBLP:conf/spaa/TsigasZ01,DBLP:conf/icdcn/Shafiei09,DBLP:conf/iceccs/ColvinG05} . We are focusing on seeing if we can implement a queue in sublinear steps in terms of $p$ or not.

\subsection{Universal constructions}
Herlihy discussed the possibility of implementing shared objects from other objects~\cite{10.1145/114005.102808}. A \textit{universal construction} is an algorithm that can implement a shared version of any given sequential object. We can implement a concurrent queue using a universal construction. Jayanti proved an $\Omega(\log p)$ lower bound on the worst-case shared-access time complexity of $p$-process universal constructions~\cite{DBLP:conf/podc/Jayanti98a}. He also introduced a construction that achieves $\textsc{O}(\log^2 p)$ shared accesses~\cite{DBLP:conf/podc/ChandraJT98}. His universal construction can be used to create any data structure, but its implementation is not practical because of using unreasonably large-sized \texttt{CAS} operations.

Ellen and Woelfel introduced an implementation of a Fetch\&Inc object with step complexity of $O(\log p)$ using $O(\log n$)-bit \texttt{LL/SC} objects, where $n$ is the number of operations~\cite{10.1007/978-3-642-41527-2_20}. Their idea has similarities to Jayanti's construction, and they represent the value of the Fetch\&Inc using the history of successful operations.

@@

\section{Our Contribution}
In this work, we design a queue with $O(\log^2 p +\log q)$ steps per operation, where $q$ is the size of the queue, avoiding the $\Omega(p)$ worst-case step complexity of existing shared queues based on linked lists or arrays. We introduce a data structure that allows processes to agree on the linearization ordering of their operations using $\textsc{O}(\log p)$ \texttt{CAS} per operation called a \textit{block tree}. Then we use the block tree as a stepping stone towards our queue algorithm.
A block tree is a tournament tree shared among $p$ processes (see Figure~\ref{fig::blocktree}). Each process has a leaf, and it appends its operations to its leaf. After that, the process tries to propagate its new operation up to the tree's root. An ordering of operations propagated up to a node is stored in that node. All processes agree on the sequence stored in the root and this is used as the linearization ordering. Our idea is similar to Jayanti and Petrovic's multi-enqueuer single-dequeuer Queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but we do not use \texttt{CAS} operations with big words and do not put a limit on the number of concurrent dequeues.

@@

We also implement methods \texttt{Get(i)}, \texttt{Index(op)} to get the $i$th propagated operation and compute the rank of a propagated operation in the linearization. \texttt{Get(i)} finds the block containing the $i$th operation in the root and then finds its sub-block recursively to reach a leaf. \texttt{Index()} is  similar but more complicated, finding super-blocks from a leaf to the root. The main challenge in each level of \texttt{Get(i)} and \texttt{Index(op)} is that it should take polylogarithmic steps with respect to $p$. After appending operations to the root, processes can find out information about the linearization ordering using \texttt{Get(i)} and \texttt{Index(op)}. Now we can implement Enqueue and Dequeue using our block tree. An \texttt{Enqueue(e)} appends an operation with input argument \texttt{e} in the block tree. To do a \texttt{Dequeue()}, first process $p$ first appends a \texttt{DEQ} operation to the tree. Then $p$ finds the rank of the \texttt{DEQ} using \texttt{Index()}, the rank of the \texttt{DEQ} and the information stored in the root about the queue $p$ computes the rank of  the \texttt{ENQ} having the answer of the \texttt{DEQ}. Finally $p$ returns the argument of that \texttt{ENQ} using \texttt{Get(i)}.
